MCMC Informed Variational Bayes

### 2017-05-07 MIVB/MIVB.tex
Nuke unneccesary shit

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}

\begin{document}

@@

Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, collectively denoted by $y_{1:T}$, with a forecaster interested in the density forecast of the unknown future value $y_{T+S+h}$, where $S \geq 1$. Assuming that a forecast of $p(y_{T+S+h} | y_{1:T})$ is available, uncertainty in the forecast can be reduced by observing and conditioning on the realisations of $y_{T+1:T+S}$ to produce $p(y_{T+S+h} | y_{1:T+S})$.

Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{1:T}$ such that $x_t$ is assumed to follow a Markovian structure. In this case $y_{T+S+h}$ depends on the latent variables $x_{T+1:T+S+h}$, and the forecast uncertainty in $y_{T+S+h}$ can be reduced by first incorporating information contained in realisations of $y_{T+1:T+S}$ to form the posterior densities of $x_{T+1:T+S+h} | x_{1:T}, \boldsymbol{\theta}, y_{1:T+S}$. With certain restrictions on the model, the additional information can be easily included via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of posterior densities of parameters and latent states must be undertaken.

One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution. MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+S+h} | y_{1:T}$ to $y_{T+S+h} | y_{1:T+S}$ after observation of $y_{T+1}, \dots, y_{T+S}$ may not be available before $y_{T+S+h}$ is observed. In this paper we assume that an MCMC that converges before time period $T + S + h$ will be required to begin, at the latest, by time $T$ and hence $y_{T}$ is the last observation that can be included in the MCMC.

Increasing the speed of estimation will allow more data up to time $T + S$ to be included in the forecast distribution, and this paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states. The the increase in computation speed and resulting increase in observations included in a forecast is contrasted with the decreased statistical accuracy associated with using an approximation to the posterior distribution. We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of previous MCMC samples to inform the design of the functional form.

\section{Variational Bayes}
\label{VB}

Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ and some approximating distribution $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.

We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T} |y_{1:T})}\right) d\boldsymbol{\theta} dx_{1:T}
\end{equation}
and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, x_{1:T})}{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T},
\end{equation}
and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).

Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
\begin{equation}
\label{GradAscent}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
\end{equation}
$\delta/\delta\boldsymbol{\lambda}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum (Robbins and Munro, 1951).
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

Ranganath, Gerrish and Blei (2014) showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
\begin{equation}
\label{ScoreDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right) 
\end{equation}
where $i = 1, \dots, N$ indicates simulations from $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)})$. 
The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.

The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and a differentiable transform $f$ such that
\begin{equation}
\label{Reparam}
\boldsymbol{\theta}, x_{1:T} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
\end{equation}
Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ results in
\begin{equation}
\label{changevar}
q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
\end{equation}
where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
\begin{align}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T}}d\boldsymbol{\theta}dx_{1:T} \nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T}). \label{reparamELBO}
\end{align}

The derivative of (\ref{reparamELBO}) can be estimated by
\begin{equation}
\label{ReparamDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
\end{equation}
where simulations of $\boldsymbol{\theta}$ and $x_{1:t}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.

\section{MCMC Informed Variational Bayes}
\label{CopulaVB}

A key decision in Variational Bayes is the choice of the functional form of the approximating distribution $q$. Much of the literature focuses on a generic form of $q$ that can flexibly be applied to many models. However the time-series nature of our application implies that MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$, which provide useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ for use in the approximate predictive density, $q(y_{T+S+h} | y_{1:T+S})$. There are two stages to this construction:
\begin{enumerate}
\item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
\item Extrapolate the form of $q$ used for $x_{1:T}$ to $x_{1:T+S}$.
\end{enumerate}
Treating the MCMC samples as `data', the functional form of $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem. Optimal marginal distributions are selected from a pool of suitable candidate for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure following Tran, Blei and Airoldi (2015). Sklar (1959) proves that any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
\begin{equation}
\label{vc1}
p(\omega_1, \dots, \omega_k) = p(\omega_1) \dots p(\omega_k) c(P(\omega_1), \dots, P(\omega_k))
\end{equation}
where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula. A vine copula is a structure to factorise this copula, increasing the flexibility given to the modeller, and is defined by a set of trees $\mathcal{V} = \{T_1, \dots, T_{k-1} \}$, where
\begin{enumerate}
\item $T_1$ has $k$ nodes corresponding to the $k$ variables, with $k-1$ edges.
\item $T_t, t = 2, \dots, p-1$ has nodes corresponding to the edges in tree $T_{t-1}$
\item Two nodes in tree $T_t$ are joined by an edge only if the associated edges in $T_{t-1}$ share a node. 
\end{enumerate}
Each edge in the vine corresponds to a bivariate copula conditioned on any variables that are shared between nodes, and the product of each of these $k(k-1)/2$ bivariate copulas forms the complete copula. Figure (\ref{fig:vinecop}) provides an example for a four dimensional $\theta$.
\begin{figure}[h]
\centering
\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
\vspace{2mm}
\caption{An example of a Vine Copula over four variables. The edges in the top tree represent unconditional bivariate copulas between the parameters on the connected nodes. The second tree edges represent the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node. The edge in the final tree represents the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.}
\label{fig:vinecop}
\end{figure}

>>>>

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{tikz}
\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{MCMC Informed Variational Bayes}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

\maketitle

\tableofcontents

@@

Consider a high frequency time series with observed values $y_{1}, \dots, y_{T+S}$, collectively denoted by $y_{1:T+S}$, with a forecaster interested in the forecast distribution of the future value $y_{T+S+h}$, where $S \geq 1$.

Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$ such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:{t-1}$, given $\boldsymbol{\theta}$ and $x_t$. One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution. MCMC can have slow convergence rates and generally does not admit parallel processing or data subsampling. In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but convergence of MCMC is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.

With $p(\boldsymbol{\theta}, x_{0:T}, y_{1:T})$ the forecast distribution of p(y_{T+S+h} | y_{T})$ is available, however incorporating the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$ is desirable, as it should provide a forecast with reduced uncertainty. The addition of $y_{T+1:T+S}$ is used to update the conditional posterior distribution of $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:T})$ to $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:{T+S})$, which can be performed with filtering techniques such as the Kalman filter with certain restrictions on the model. If these restrictions are not met, or the filtering technique is too computationally intensive to perform before $T+S+h$, then the forecaster is unable to obtain $p(y_{T+S+h} | y_{T+S})$.

An alternative to MCMC is Variational Bayes (VB), which approximates the posterior distribution with a tractable alternative denoted by $q$, optimising the parameters of this approximation so that the approximation is close by some metric to the actual posterior. A VB optimisation algorithm is often orders of magnitude faster than MCMC with the application of parallel processing and data subsampling available, so the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, so the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be obtained.

The use of VB introduces an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error. This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), that emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.

\begin{itemise}
\item Section Two: Variational Bayes, Optimisation and ADVI
\item Section Three: MCMC Informed Variational Bayes
\item Section Four: Simulation results for DLM, SVM
\item Section Five: Empirical SVM
\item Section Six: Discussion
\end{itemise}

\section{Variational Bayes}
\label{VB}
Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$ and some approximating distribution $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function. The notation for $q$ supresses the dependence of the distribution on $y_{1:T+S}$ as $\boldsymbol{\lambda}$ is typically a function of $y_{1:T+S}$.

This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
\end{equation}
and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \ln(p(y_{1:T+S})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T+S}))$.  $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{1:T+S})}{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T+S},
\end{equation}
and as $\ln(p(y_{1:T+S}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).

\subsection{ELBO Optimisation}
Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where the following update step is iteratively applied until (\ref{ELBO}) converges within some pre-specified tolerance:
\begin{equation}
\label{GradAscent}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S}).
\end{equation}
$\delta/\delta\boldsymbol{\lambda}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum (Robbins and Munro, 1951).
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

Ranganath, Gerrish and Blei (2014) showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
\begin{equation}
\label{ScoreDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T+S}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right) 
\end{equation}
where $i = 1, \dots, N$ indicates simulations from $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)})$. 
The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.

The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and a differentiable transform $f$ such that
\begin{equation}
\label{Reparam}
\boldsymbol{\theta}, x_{1:T+S} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
\end{equation}
Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ results in
\begin{equation}
\label{changevar}
q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
\end{equation}
where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T+S}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
\begin{align}
\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
\end{align}

The derivative of (\ref{reparamELBO}) can be estimated by
\begin{equation}
\label{ReparamDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
\end{equation}
where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.

\subsection{$q$ Functional Form}
A key decision in Variational Bayes is the choice of the functional form of the approximating distribution $q$. Much of the literature focuses on a `black box' functional form, where the same $q$ can be applied to many models. One popular black box method is Automatic Differentation Variational Bayes (ADVI, Tran?), which first transforms each parameter $\theta$ and latent variable $x_t$ to the real line then selects a multivariate normal $q$ to approximate the posterior distribution of the transformed parameters with either a diagonal or non-diagonal covariance matrix.

For illustration, consider a common latent variable time series model, the Dynamic Linear Model, described by (\ref{DLM:Measure}) and (\ref{DLM:Transition}).

\begin{align}
y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}
where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables. The priors used are
\begin{align}
\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
\phi &\sim \mathcal{U}(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}
where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution. Conjugate priors are chosen for $\gamma, \sigma^2_y$ and $\sigma^2_x$, while the uniform prior for $\phi$ constrains $p(\phi | y_{1:T+S})$ to have zero probability mass outside of the AR(1) stationary region. $\sigma^2_y$ and $\sigma^2_x$ are log-transformed such that $\theta^* = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x), x_{0:T}\} \in \mathbb{R}^{T+5}$. $100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$. For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5,000$ and $10,000$ as this has a complexity of $O(T^2)$. MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half. The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$. Section \label{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.
